---
title: "MERMAID Image Classification Results"
subtitle: "Visualizations to show model perfomance"
author: "Iain R. Caldwell"
date: 06/03/2025
format: 
  html: #for website
    embed-resources: true
editor: visual
code-fold: true
code-summary: "Show the code"
toc: true
title-block-banner: "#f0f3f5"
title-block-banner-color: "black"
include-after-body: "footer.html"
---

## Context - Visualizing MERMAID image classification performance

This code visualizes model performance results from the MERMAID image classification model.

## Loading package libraries and setting parameters

The following libraries and parameters are used throughout the code.

```{r}
#| label: Load package libraries and set parameters
#| warning: false
rm(list = ls()) #remove past stored objects
options(scipen = 999) #turn off scientific notation

#Package libraries
library(readxl)
library(tidyverse)
library(ggplot2)
library(mermaidr)
library(ggtext)
library(xfun)
```

## Loading results files

```{r}
#| label: Load data
#| warning: false
### Report with the overall performance
overallClassReportTBL <- 
  read_excel(path = "../data/classifier_metrics.xlsx",
             sheet = "Metrics")

### Report with the performance per label (UUIDs)
labelClassReportTBL <-
  read_excel(path = "../data/classifier_metrics.xlsx",
             sheet = "Classification_Report")

### Label mapping --> UUID to character
labelMapTBL <- read_csv(file = "../data/LabelMap.csv")

### MERMAID benthic attributes
benthicAttTBL <- mermaid_get_reference(reference = "benthicattributes") %>% 
  select(name, parent) %>% 
  rename(ba = name)

```

## Prepare data

```{r}
#| label: Prepare data
overallClassReportTBL <- overallClassReportTBL %>% 
  filter(...1 %in% c("precision", "recall", "f1_score")) %>% 
  pivot_wider(names_from = ...1,
              values_from = `0.0`) %>% 
  mutate(CoralFocus3Label = "<b>Overall</b>",
         ba = "Overall",
         tlc = "Overall") %>% 
  rename(`f1-score` = f1_score)

labelClassReportTBL <- labelClassReportTBL %>% 
  rename(label_id = ...1) %>% 
  filter(!label_id %in% c("accuracy", "weighted avg", "macro avg")) %>% 
  left_join(labelMapTBL, by = "label_id") %>% 
  select(CoralFocus3Label, `f1-score`, precision, recall)

#### Get the top level categories for each of the labels
## Extract just the benthic attribute from the label
labelClassReportTBL <- labelClassReportTBL %>% 
  mutate(ba = gsub(pattern = paste(" -",
                                   c("Branching",
                                     "Foliose",
                                     "Encrusting",
                                     "Plates or tables",
                                     "Massive",
                                     "Digitate"),
                                   collapse = "|"),
                   replacement = "",
                   x = CoralFocus3Label))

### Get all the unique benthic attributes and assign them to the top level categories
uniqueBaParentTBL <- labelClassReportTBL %>% 
  select(ba) %>% 
  distinct() 

# Function to find the top-level category
find_top_level <- function(ba, lookup_table) {
  parent <- lookup_table$parent[lookup_table$ba == ba]
  if (is.na(parent)) {
    return(ba) # If no parent, the current ba is the top-level category
  } else {
    # Recursively find the top-level category
    return(find_top_level(parent, lookup_table))
  }
}

uniqueBaTlcTBL <- uniqueBaParentTBL %>%
  rowwise() %>%
  mutate(tlc = find_top_level(ba, benthicAttTBL)) %>%
  ungroup()

labelClassReportTBL <- labelClassReportTBL %>% 
  left_join(uniqueBaTlcTBL, by = "ba")

##Get the maximum f1 score by tlc to order tlc as a factor
f1ScoreByTlcTBL <- labelClassReportTBL %>% 
  group_by(tlc) %>% 
  summarise(max_f1_score = max(`f1-score`)) %>%
  ungroup() %>% 
  arrange(desc(max_f1_score))

allClassReportTBL <- bind_rows(labelClassReportTBL, overallClassReportTBL) 

```

## Plot the results - horizontal barplots by label

```{r}
#| label: Horizonal barplot by label
#| fig-height: 10
#| fig-width: 10

# Reshape data for faceted plotting
longAllClassReportTBL <- allClassReportTBL %>%
  pivot_longer(cols = c(precision, recall, `f1-score`),
               names_to = "Metric", values_to = "Score") %>% 
  mutate(Group = ifelse(CoralFocus3Label == "<b>Overall</b>", "Overall", "Label"))

# Order labels by F1-score 
label_order <- labelClassReportTBL %>%
  arrange(`f1-score`) %>%
  pull(CoralFocus3Label)

tlc_order <- f1ScoreByTlcTBL %>% 
  pull(tlc)

longAllClassReportTBL <- longAllClassReportTBL %>%
  mutate(CoralFocus3Label = factor(CoralFocus3Label,
                                   levels = c(label_order, "<b>Overall</b>")),
         Metric = factor(Metric, levels = c("f1-score", "precision", "recall")),
         Group = factor(Group, levels = c("Overall", "Label")),
         tlc = factor(tlc, levels = c("Overall", tlc_order)))

# Create a plot with group as the row facets
ggplot(longAllClassReportTBL,
       aes(x = Score, y = CoralFocus3Label, fill = Score)) +
  facet_grid(Group ~ Metric, scales = "free_y", space = "free_y") +
  geom_col(alpha = 0.75) +
  geom_vline(xintercept = c(0, 0.25, 0.5, 0.75, 1), linetype = "dotted") +
  scale_fill_viridis_c(option = "D", direction = -1) +
  scale_x_continuous(breaks = c(0, 0.25, 0.5, 0.75, 1),
                     labels = c(0, 0.25, 0.5, 0.75, 1)) +
  labs(x = "Score",
       y = NULL,
       title = "MERMAID Classification Model Performance",
       subtitle = "Overall results and by label") +
  theme_minimal(base_size = 14) +
  theme(
    strip.text = element_text(face = "bold", size = 14),
    strip.text.y.right = element_blank(),
    axis.text.x = element_text(size = 10),
    axis.ticks.x = element_line(),
    axis.title.x = element_text(size = 12, face = "bold", margin = margin(t = 10)),
    axis.text.y = element_markdown(size = 10),
    panel.spacing.x = unit(0.75, "lines"),
    legend.position = "none"
  )

tlcColors <- c(Overall = "black",
               `Hard coral` = "#498FC9",
               Macroalgae = "#B2B002",
               Sand = "#C1B180",
               `Soft coral` = "#9BE5FA",
               Rubble = "#F5F7AF",
               `Other invertebrates` = "#A6A6A6",
               Cyanobacteria = "#860E00",
               `Bare substrate` = "#F2F3F3",
               Seagrass = "#4D4D4D",
               `Crustose coralline algae` = "#FBD7D5",
               `Turf algae` = "#D8EEA8")

tlcLabels <- c(Overall = "",
               `Hard coral` = "Hard coral",
               Macroalgae = "",
               Sand = "",
               `Soft coral` = "Soft coral",
               Rubble = "",
               `Other invertebrates` = "Other invertebrates",
               Cyanobacteria = "",
               `Bare substrate` = "",
               Seagrass = "",
               `Crustose coralline algae` = "",
               `Turf algae` = "")

# Create a plot with tlc as the row facets
ggplot(longAllClassReportTBL,
       aes(x = Score, y = CoralFocus3Label, fill = tlc)) +
  facet_grid(tlc ~ Metric, scales = "free_y", space = "free_y",
             labeller = labeller(tlc = tlcLabels)) +
  geom_col(alpha = 0.75) +
  geom_vline(xintercept = c(0, 0.25, 0.5, 0.75, 1), linetype = "dotted") +
  scale_x_continuous(breaks = c(0, 0.25, 0.5, 0.75, 1),
                     labels = c(0, 0.25, 0.5, 0.75, 1)) +
  scale_fill_manual(values = tlcColors) +
  labs(x = "Score",
       y = NULL,
       title = "MERMAID Classification Model Performance",
       subtitle = "Overall results and by label") +
  theme_minimal(base_size = 14) +
  theme(
    strip.text = element_text(face = "bold", size = 12),
    #strip.text.y.right = element_blank(),
    strip.text.y.right = element_text(angle = 0, hjust = 0),
    axis.text.x = element_text(size = 10),
    axis.ticks.x = element_line(),
    axis.title.x = element_text(size = 12, face = "bold", margin = margin(t = 10)),
    axis.text.y = element_markdown(size = 10),
    panel.spacing.x = unit(0.25, "lines"),
    legend.position = "none"
  )
```
